---
output: pdf_document
geometry: margin=0.7in
title: Prediction Model for Patients' Length of Stay in Emergency Room
author: Yifan Zhou
---

```{r results='hide',message=F,warning=F, echo=F}
options(digits=2)
library(ggplot2)
library(splines)
library(plyr)
library(lmtest)
library(grid)
library(gridExtra)
library(mice)
library(MASS)
library(xtable)
library(boot)
```

```{r results='hide',message=F,warning=F, echo=F}
#explore data 
data<-read.csv("2016_comp_dap.csv")
#nrow(data): 50764
#sum(is.na(data$key)): 9642 n/a in key

data$week<-data$weekd-1
data$week<-ifelse(data$week==0,7,data$week)
data$age_sp1<-ifelse(data$age<=25,0,data$age-25)
data$age_sp2<-ifelse(data$age<=85,0,data$age-85)
data$weekday<-ifelse(data$week<=5,1,0)
data$acu23<-ifelse(data$acuity==2|data$acuity==3,1,0)
data$prenoon<-ifelse(data$tod<12,1,0)
data$key<-as.character(data$key)
data$key<-ifelse(data$key=="a & d","ad",data$key)
data$key<-ifelse(data$key=="chest pain","chestpain",data$key)

#key
#countk<-count(data,'key')
#countk<-as.data.frame(countk)
#head(countk)
#keyname<-countk$key
#meanlos<-rep(NA,nrow(countk))
#medianlos<-rep(NA,nrow(countk))
#for (i in 1:nrow(countk)){
#  meanlos[i]<-mean(subset(data,data$key==keyname[i])$los)
#  medianlos[i]<-median(subset(data,data$key==keyname[i])$los)
#}
#countk$meanlos<-meanlos
#countk$type<-ceiling(countk$meanlos/100)
#countk$type<-ifelse(countk$type>6,6,countk$type)
#write.csv(countk,file="key.csv")
#for (i in 1:nrow(data)){
#  data$type[i]<-countk$type[which(countk$key %in% data$key[i])]
#}
#data<-data[,-7]
#imdata <-mice(data, m=5,printFlag = FALSE)
#imm<-matrix(NA,nrow=nrow(data),ncol=5)
#for(i in 1:5){
#  imm[,i]<-complete(imdata,i)$type
#}
#imtype<-round(apply(imm,1,mean))
#data$type<-imtype
#data<-data[,-c(1,2,3,8)]
#data$gender<-ifelse(data$gender=="U",NA,data$gender)  #gender==1 female
#data$gender<-ifelse(data$gender==2,0,data$gender)
#data<-data[complete.cases(data),]
#write.csv(data1,"newdata.csv")
#write.csv(countk,"keycategory.csv")
```


```{r results='hide',message=F,warning=F, echo=F}

# tod ??? ns(tod,7) change to prenoon
#plot(data$tod,data$los,cex=0.4,pch=16,col=rgb(0.5,0.5,0.5,0.4))
#lines(ksmooth(data$tod,data$los,kernel="normal",bandwidth = 3),col=2,lwd=2)
#fig.tod <- ggplot(data, aes(x=factor(tod), y=los)) + theme(legend.position="bottom") +geom_point(size=0.1,alpha=0.4)+theme_bw()+geom_boxplot()+ylim(0,600)
#fig.tod

# acuity ?  factor
#fig.acu <- ggplot(data, aes(x=factor(acu), y=los)) + theme(legend.position="bottom") +geom_point(size=0.1,alpha=0.4)+theme_bw()+geom_boxplot()
#plot(data$acuity,data$los,cex=0.4,pch=16,col=rgb(0.5,0.5,0.5,0.4),ylim=c(0,800))
#lines(ksmooth(data$acuity,data$los,kernel="normal",bandwidth = 1))
 #acuity:prenoon
#fig.acu_prenoon <- ggplot(data, aes(x=factor(acu), y=los,group=prenoon,colour=factor(prenoon))) + theme(legend.position="bottom") +geom_point(size=0.1,alpha=0.4)+theme_bw()+stat_smooth(method="loess",alpha=0,size=0.5,span=2)
#fig.acu_prenoon
 #acuity:weekday **
fig.acu_weekday <- ggplot(data, aes(x=factor(acuity), y=los,group=weekday,colour=factor(weekday))) + ylim(0,2000) +geom_jitter(size=0.1,alpha=0.2)+theme_bw()+stat_smooth(alpha=0,size=1.2)+scale_color_discrete(name="Day of the week",labels=c("weekend","weekday"))+theme(legend.position="bottom")+theme(text = element_text(size=7,face="bold"))+ggtitle("Fig 2")+xlab("acu")
#fig.acu_weekday

#gender-acuity ** acuity=1,2,5
#fig.acu_gender <- ggplot(data, aes(x=acu, y=los,group=gender,colour=factor(gender))) + theme(legend.position="bottom") +geom_point(size=0.1,alpha=0.4)+theme_bw()+stat_smooth(method="loess",alpha=0,size=0.5,span=2)
#fig.acu_gender

# week *** F6,Sa7,Su1 small ???"natural spline"" or "dummy variable", M is large
#fig.week <- ggplot(data, aes(factor(week), y=los)) + theme(legend.position="bottom") +geom_point(size=0.1,alpha=0.4)+theme_bw()+geom_boxplot()
#fig.week
#plot(data$week,data$los,cex=0.4,pch=16,col=rgb(0.5,0.5,0.5,0.4),ylim=c(0,800))
#lines(ksmooth(data$week,data$los,kernel="normal",bandwidth = 1),col=2,lwd=2)

# age knots= 30,60 linear spline
#fig.age <- ggplot(data, aes(age, los)) + theme(legend.position="bottom") +geom_point(size=0.1,alpha=0.4)+stat_smooth(method="loess",alpha=0,size=1.5,span=2)

 #age:acuity 
#fig.age_acuity <- ggplot(data, aes(x=age, y=los,group=acuity,colour=factor(acuity))) + theme(legend.position="bottom")+ylim(0,1000) +geom_point(size=0.1,alpha=0.4)+theme_bw()+stat_smooth(method="loess",alpha=0,size=1,span=1)

#age:weekday *
fig.age_weekday <- ggplot(data, aes(x=age, y=los,group=weekday,colour=factor(weekday))) +ylim(0,1000) +geom_jitter(size=0.1,alpha=0.2)+theme_bw()+stat_smooth(alpha=0,size=1.5)+scale_color_discrete(name="Day of week",labels=c("weekend","weekday"))+theme(legend.position="bottom")+theme(text = element_text(size=7,face="bold"))
#fig.age_weekday

 #age:prenoon **
fig.age_prenoon <- ggplot(data, aes(x=age, y=los,group=prenoon,colour=factor(prenoon)))+ylim(0,1000) +geom_jitter(size=0.1,alpha=0.2)+theme_bw()+stat_smooth(alpha=0,size=1.2)+ggtitle("Fig 1")+theme(text = element_text(size=7,face="bold"))+scale_color_discrete(name="Hour of day",labels=c("[12PM-11:59PM]","12AM-11:59AM"))+theme(legend.position="bottom")
#fig.age_prenoon

 #age:gender  ***
fig.age_gender <- ggplot(data, aes(x=age, y=los,group=gender,colour=factor(gender)))+geom_jitter(size=0.1,alpha=0.2)+theme_bw()+stat_smooth(alpha=0,size=1.5)+scale_color_discrete(name="Gender",labels=c("Male","Female"))+theme(legend.position="bottom")+ylim(0,1000)+theme(text = element_text(size=7,face="bold"))
#fig.age_gender
 
 #gender-week
#fig.week_gender <- ggplot(data, aes(x=week, y=los,group=gender,colour=factor(gender))) + theme(legend.position="bottom") +geom_point(size=0.1,alpha=0.4)+theme_bw()+stat_smooth(method="loess",alpha=0,size=1,span=1)
 #gender-acuity ** acuity=1,2,5
#fig.acuity_gender <- ggplot(data, aes(x=acuity, y=los,group=gender,colour=factor(gender))) + theme(legend.position="bottom") +geom_point(size=0.1,alpha=0.4)+theme_bw()+stat_smooth(method="loess",alpha=0,size=0.5,span=2)

```

## I. Research Question

The research interest is to estimate the length of emergency room stay until discharge or hospital admission (*los*) given patients' information when they first arriave at the ED and give the range of plausible estimated values. 

## II. Data Transformation 

To explore the relationship between 7 predictors and outcome, we draw several plots then decide to do some data transformation and create new variables: *weekday* is a dummy variable indicating whether the patient arrive on weekday; *prenoon* is a dummy variable indicating whether the patient arrive before the noon (*tod*=[12AM-11:59AM]); *acu23* is a dummy variable indicating *acuity* is level "2" or "3"; For the presenting symptoms category, we classify the 163 symptoms into 6 types (*type* variable) according to the average *los* for patients with each symptom (see Supplement). There are 9642 missing values (19\% of the data set) in *type* variable, since dropping them will lose many information, we decide use multivariate imputation method to impute the missing *type* variable. There are only 329 (0.65\% of the data set) patients with unknown gender, then we just dropped those data.

## III. Model

Since the outcome variable (length of stay) is recorded in minutes, it can be treated as count data. While the data are highly over-dispered (the variance of *los* is 400 times over its mean), we decide to use the negative binomial regression model with log link function.

To determine the full model, we plot the outcome variable *los* against *age* and stratified by other categorical variables, we also plot the outcome variable against different acuity level (*acuity*) and stratified by *weekday*.

```{r results='hide',message=F,warning=F, echo=F,fig.height=3.5, dev="png", dpi=200}
par(mar=c(2,2,2,0),mgp=c(0.8,0.1,0),fin=c(1,1.5))
layout(matrix(c(1,2),nrow=1,ncol=2,byrow=T))
#Fig 1
prndata<-data[data$prenoon==1,]
afndata<-data[data$prenoon!=1,]
plot(jitter(prndata$age),prndata$los,cex=0.3,pch=16,col=rgb(0.1,0.5,0.8,0.2),
     main="Fig 1",xlab="age",ylab="los",ylim=c(0,1000),cex.lab=0.5,yaxt="n",xaxt="n")
axis(1,cex.axis=0.5,tck=-.01)
axis(2,cex.axis=0.5,tck=-.01)
points(jitter(afndata$age),afndata$los,cex=0.3,pch=16,col=rgb(0.8,0.5,0.1,0.2))
lines(loess.smooth(prndata$age,prndata$los,span=0.2),col=rgb(0.1,0.5,0.8),lwd=3)
lines(loess.smooth(afndata$age,afndata$los,span=0.2),col=rgb(0.8,0.5,0.1),lwd=3)
legend("topright",col=c(rgb(0.1,0.5,0.8),rgb(0.8,0.5,0.1)),legend = c("12AM-11:59AM","12PM-11:59PM"),title = "Hour of day",pch=16,cex=0.5)

#Fig 2
weekdata<-data[data$weekday==1,]
wenddata<-data[data$weekday!=1,]
plot(jitter(weekdata$acuity),weekdata$los,cex=0.3,pch=16,col=rgb(0.1,0.5,0.8,0.2),
     main="Fig 2",xlab="acuity",ylab="los",ylim=c(0,2000),cex.lab=0.5,yaxt="n",xaxt="n")
axis(1,cex.axis=0.5,tck=-.01)
axis(2,cex.axis=0.5,tck=-.01)
points(jitter(wenddata$acuity),wenddata$los,cex=0.3,pch=16,col=rgb(0.8,0.5,0.1,0.2))
lines(ksmooth(weekdata$acuity,weekdata$los,kernel="normal",bandwidth = 1),col=rgb(0.1,0.5,0.8),lwd=3)
lines(ksmooth(wenddata$acuity,wenddata$los,kernel="normal",bandwidth = 1),col=rgb(0.8,0.5,0.1),lwd=3)
legend("topright",col=c(rgb(0.1,0.5,0.8),rgb(0.8,0.5,0.1)),legend = c("weekday","weekend"),title = "Day of week",pch=16,cex=0.5)
data<-read.csv("newdata.csv")
data<-data[,-1]
for (i in 2:6){
  data[,(10+i)]<-ifelse(data$type==i,1,0)
}
colnames(data)[12:16]<-c("t2","t3","t4","t5","t6")
```

```{r results='hide',message=F,warning=F, echo=F,fig.height=2.6, dev="png", dpi=200}
fit1<-glm.nb(los~(age+age_sp1+age_sp2)*(acu23+gender+weekday+prenoon)+acu23*(gender+weekday+prenoon)+ gender*(weekday+prenoon)+ weekday*prenoon + factor(type) , data=data)
#summary(fit1)

fit0<-glm.nb(los~age*(acu23+gender+weekday+prenoon)+acu23*(gender+weekday+prenoon)+ gender*(weekday+prenoon)+ weekday*prenoon + factor(type) ,data=data)
lrtage<-lrtest(fit0,fit1)
page<-lrtage$`Pr(>Chisq)`[2]

#gender:weekday
#fit2<-glm.nb(los~(age+age_sp1+age_sp2)*(acu23+gender+weekday+prenoon)+acu23*(gender+weekday+prenoon)+ gender*prenoon+ weekday*prenoon + factor(type) , data=data)
#lrtest(fit2,fit1)
#summary(fit2)

#age:weekday
#fit3<-glm.nb(los~age*(acu23+gender+prenoon)+(age_sp1+age_sp2)*(acu23+gender+weekday+prenoon)+acu23*(gender+weekday+prenoon)+ gender*prenoon+ weekday*prenoon + factor(type), data=data)
#lrtest(fit3,fit2)
#summary(fit3)
#acu23:gender
#fit4<-glm.nb(los~age*(acu23+gender+prenoon)+(age_sp1+age_sp2)*(acu23+gender+weekday+prenoon)+acu23*(weekday+prenoon)+ gender*prenoon+ weekday*prenoon + factor(type), data=data)
#lrtest(fit4,fit3)
#summary(fit4)
#age_sp2:prenoon
#fit5<-glm.nb(los~age*(acu23+gender+prenoon)+age_sp1*(acu23+gender+weekday+prenoon)+age_sp2*(acu23+gender+weekday)+acu23*(weekday+prenoon)+ gender*prenoon+ weekday*prenoon + factor(type), data=data)
#lrtest(fit5,fit4)
#summary(fit5)
#gender:prenoon
#fit6<-glm.nb(los~age*(acu23+gender+prenoon)+age_sp1*(acu23+gender+weekday+prenoon)+age_sp2*(acu23+gender+weekday)+acu23*(weekday+prenoon)+ weekday*prenoon + factor(type), data=data)
#lrtest(fit6,fit5)
#summary(fit6)
#age_sp1:gender
#fit7<-glm.nb(los~age*(acu23+gender+prenoon)+age_sp1*(acu23+weekday+prenoon)+age_sp2*(acu23+gender+weekday)+acu23*(weekday+prenoon)+ weekday*prenoon + factor(type), data=data)
#lrtest(fit7,fit6)
#summary(fit7)
#age:gender
#fit8<-glm.nb(los~age*(acu23+prenoon)+age_sp1*(acu23+weekday+prenoon)+age_sp2*(acu23+gender+weekday)+acu23*(weekday+prenoon)+ weekday*prenoon + factor(type), data=data)
#lrtest(fit8,fit7)
#summary(fit8)
```

```{r results='hide',message=F,warning=F, echo=F,fig.height=2.6, dev="png", dpi=200}
#acu23:prenoon
fit9<-glm.nb(los~age*(acu23+prenoon)+age_sp1*(acu23+weekday+prenoon)+age_sp2*(acu23+gender+weekday)+acu23*weekday+ weekday*prenoon + factor(type), data=data)
#lrtest(fit9,fit8)
#summary(fit9)

fit<-fit9
```

Fig 1 showes that the length of stay has an non-linear relationship with age, therefore we apply linear spline on *age* with knots at 25 and 85. The result of the likelihood ratio test comparing with the model without linear spline ($p=`r page`$) shows that linear spline significantly improves the model fitting. 
Fig 1 also shows that patients arrive before noon (blue curve) have longer length of stay, and the difference between *los* before and after noon changes with *age*. Therefore, we include the interaction term of *age* and *prenoon*. 
In the same way, we include the interaction terms of *age* and *weekday*, *age* and *gender*. 
Fig 2 shows there is significant relationship between length of stay and acuity level, and it may be modified by *weekday*. Therefore we include the interaction term of acuity level and weekday. Also, we add the indicators of 6 symptoms types (dummy variable of *type*) as the covariates into the model.

Then we fit a full model and performe several likelihood ratio tests to drop non-significant terms and decide the fianl model.

## IV. Assumptions and Model Checking

First, we check for the independence assumption by checking the correlation between two patients' standardized residuals. The correlation plot  shows there is no significant correlation between two patients.
Then, we check if average standardized residual equals to 0 at each level of predicted values by plotting the average standarized residual in different predicated value intervals, and the result shows the final model is a good fit.
We also check for the highly influential points by dropping points using the criteria "dffits>0.17" and 59 points are dropped. Then we refit the same model and compare the coefficients and predicted values before and after dropping. Result shows that there are few highly influential points and the model is robust enough. The result of 10-fold cross-validation shows the model's prediction performance is pretty good and there is no overfitting. Results are shown in Supplement.

## V. Results and Interpretation

By using the final model fitted in III, we could predict the length of stay in the ED and calculate the 95\% confidence interval of the predicted values.

Fig 3 shows the predicted length of stay with 95\% confidence interval for female patients who arrive on weekday during 12AM-11:59AM with acuity level 2 or 3, and the results are stratified by 6 symptom types indicated by color. Through Fig 3, we could see the predicted length of stay gets longer for patients from type 1 to type 6, when controlled for other predictors. This phenomenon is as expected since those types are generated by the average length of stay. The confidence intervals in Fig 3 show that for patients younger than 25 years old, the variance of predicted values decreases with age, while for patients older than 25 years old the variance will increase with age. And the variance of predicated value for patients in type 1, 2, 3 will be smaller than the variance for patients in type 4, 5, 6. The results for male patients and other arrival time and acuity levels are similar to those conclutions.


```{r message=F,warning=F, echo=F,fig.height=3.5, dev="png", dpi=200}
# ci for yhat
#pred<-predict.glm(fit,type="response")
#resi<-data$los-pred
#presi<-(data$los-pred)/sqrt(pred+pred^2*fit$theta)

#----------------------------------------------type:gender
age1<-rep(NA,105*6)
for(i in 1:105){
  age1[((i-1)*6+1):(i*6)]<-rep(i,6)
}
result<-data.frame()
for(i in 1:length(age1)){
  result[i,1]<-age1[i]
  result[i,2]<-ifelse(age1[i]<=25,0,age1[i]-25)
  result[i,3]<-ifelse(age1[i]<=85,0,age1[i]-85)
}

result[,4]<-rep(c(1:6),105)
result[,5:9]<-model.matrix(~as.factor(result[,4]))[,2:6]
colnames(result)<-c("age","age_sp1","age_sp2","type","type2","type3","type4","type5","type6")
result$weekday<-rep(1,nrow(result))
result$prenoon<-rep(1,nrow(result))
result$acu23<-rep(1,nrow(result))
result$gender<-rep(1,nrow(result))  ## male
#result0<-result  #
#result0$weekday<-0
#result1<-result  #
#result<-rbind(result0,result1)

inter<-rep(1,nrow(result))
X1<-data.frame(inter, result$age, result$acu23, result$prenoon, result$age_sp1, result$weekday, result$age_sp2, result$gender)
X2<-data.frame(result[,5:9])
X3<-data.frame(result$age*result$acu23, result$age*result$prenoon, result$age_sp1*result$acu23, result$age_sp1*result$weekday, result$age_sp1*result$prenoon, result$age_sp2*result$acu23, result$age_sp2*result$gender,result$age_sp2*result$weekday , result$weekday*result$acu23, result$prenoon*result$weekday)
X<-cbind(X1,X2,X3)
X<-as.matrix(X)
B<-as.matrix(coefficients(fit))
XB<-X%*%B
varxb<-X%*%vcov(fit)%*%t(X)
sexb<-sqrt(diag(varxb))
ci<-data.frame("lowerxb"=XB-2*sexb,"upperxb"=XB+2*sexb,"lowery"=exp(XB-2*sexb),"uppery"=exp(XB+2*sexb),"yhat"=exp(XB), "se"=sexb)
output<-cbind(result,ci)

#output.1<-output[output$gender==0,]
#output.2<-output[output$gender==1,]


fig.ci_type<-ggplot(output,aes(age))+geom_line(aes(y=yhat,group=type,colour=factor(type)))+geom_ribbon(aes(ymin=lowery,ymax=uppery,group=type,fill=factor(type)),alpha=0.4)+ggtitle("Fig 3: Predicted values with 95% CI")+theme_bw()+ylab("estimated length of stay")+theme(legend.position="bottom")+theme(text = element_text(size=7,face="bold"))

pred<-predict.glm(fit,type="response")
resi<-data$los-pred
presi<-(data$los-pred)/sqrt(pred+pred^2*fit$theta)
data$pred<-pred
data$presi<-presi

x<-seq(1,105)
prob<-function(data,x){
  m<-rep(NA,length(x))
  mt1<-rep(NA,length(x))
  mt0<-rep(NA,length(x))
  for(i in 1:length(x)){
    mt1[i]<-mean(data$los[data$age==x[i] &  data$weekday==1 ])
    mt0[i]<-mean(data$los[data$age==x[i] &  data$weekday==0 ])
    m[i]<-mean(data$los[data$age==x[i] ])
  }
  obser1<-data.frame("age"=x,"meanlos"=m,"weekday"=2)
  obser2<-data.frame("age"=x,"meanlos"=mt1,"weekday"=1)
  obser3<-data.frame("age"=x,"meanlos"=mt0,"weekday"=0)
  obser<-rbind(obser1,obser2,obser3)
  return(obser)
}
obser<-prob(data,x)

fig.weekday<-ggplot(data,aes(age))+stat_smooth(aes(y=pred,group=weekday,colour=factor(weekday)),alpha=0)+
stat_smooth(aes(y=pred,group=NA),col="sky blue",alpha=0)+geom_point(data=obser,aes(age,meanlos,group=weekday,color=factor(weekday)),size=0.5)+
ggtitle("Fig 4: Predicted curves and observed values")+theme_bw()+ylab("estimated length of stay")+theme(legend.position="bottom")+theme(text = element_text(size=7,face="bold"))+
scale_color_discrete(name="Day",labels=c("Weekend","Weekday","All data"))

grid.arrange(fig.ci_type,fig.weekday,nrow=1,bottom = textGrob("Fig 3 is for female patients arrive on weekday during 12AM-11:59AM and have acuity level 2 or 3; Fig 4 is for all patients", gp = gpar(fontface=3, fontsize=8),hjust=1, x=1))

calci<-function(age,type,gender,weekday,prenoon,acu23){
  age_sp1<-ifelse(age<=25,0,age-25)
  age_sp2<-ifelse(age<=85,0,age-85)
  X1<-data.frame(1, age, acu23, prenoon, age_sp1, weekday, age_sp2, gender)
  X2<-data.frame(result[type,5:9]) #type
  X3<-data.frame(age*acu23, age*prenoon, age_sp1*acu23, age_sp1*weekday, age_sp1*prenoon, age_sp2*acu23, age_sp2*gender, age_sp2*weekday, weekday*acu23, prenoon*weekday)
  X<-as.matrix(cbind(X1,X2,X3))
  B<-as.matrix(coefficients(fit))
  XB<-X%*%B
  varxb<-X%*%vcov(fit)%*%t(X)
  sexb<-sqrt(diag(varxb))
  ci<-data.frame("y"=exp(XB),"l"=exp(XB-2*sexb),"u"=exp(XB+2*sexb))
  return(ci)
}

#acu23=1
# weekday=1          #    weekday=0
#prenoon=1 prenoon=0 # prenoon=1 prenoon=0
table<-matrix(NA,ncol=12,nrow=10)
for(i in 1:10){
  table[i,]<-as.matrix(cbind(calci(i*10,4,1,1,1,1), calci(i*10,4,1,1,0,1), calci(i*10,4,1,0,1,1), calci(i*10,4,1,0,0,1)))[1,]
}
table1<-as.data.frame(table)
#xtable(table1)

#acu23=0
table<-matrix(NA,ncol=12,nrow=10)
for(i in 1:10){
  table[i,]<-as.matrix(cbind(calci(i*10,4,1,1,1,0), calci(i*10,4,1,1,0,0), calci(i*10,4,1,0,1,0), calci(i*10,4,1,0,0,0)))[1,]
}
table2<-as.data.frame(table)
#xtable(table2)

```


In Fig 4, the points are the observed average length of stay for each age and the lines are smooth curves of predicted values, and color indicates whether patients arrive on weekday or weekend. This figure shows the final model is a good fit for the data and *weekday* is an important predictor in predicting the lenght of stay: Patients arrive on weekday will have longer predicted length of stay compared with patients in same condition but arrive on weekend.

Fig 3 and Fig 4 both show the relationship between age and predicted length of stay. For people in age group 0-85 years old, older patients have longer predicted length of stay when keep other predictors as the same; For people in age group 85-105 years old, older patients will have shorter predicted values when keep other predictors as the same.

To specify, we display the prediction results for female patients in a specific type group (type 4) shown by the two tables above. Table 1 and Table 2 summary the predicted values with 95\% confidence interval by age, acuity level, arrival day and time. 

\begin{table}[]
\centering
\caption{Predicted values with 95\% CI for female patients with acuity level 2 and 3 (symptom type=4)}
\label{my-label}
\begin{tabular}{|l|ll|ll|ll|ll|}
\hline
\multicolumn{1}{|c|}{\textbf{Day}}  & \multicolumn{4}{c|}{\textbf{Weekday}}                                                               & \multicolumn{4}{c|}{\textbf{Weekend}}                                                               \\ \hline
\multicolumn{1}{|c|}{\textbf{Time}} & \multicolumn{2}{c|}{\textbf{{[}12AM-11:59AM{]}}} & \multicolumn{2}{c|}{\textbf{{[}12PM-11:59PM{]}}} & \multicolumn{2}{c|}{\textbf{{[}12AM-11:59AM{]}}} & \multicolumn{2}{c|}{\textbf{{[}12PM-11:59PM{]}}} \\ \hline
Age                                 & $\widehat{los}$               & 95\% CI              & $\widehat{los}$               & 95\% CI              & $\widehat{los}$               & 95\% CI              & $\widehat{los}$               & 95\% CI              \\ \hline  
  10 & 257.19 & (235.55 , 280.83) & 220.99 & (206.76 , 236.20) & 218.31 & (199.88 , 238.43) & 173.44 & (162.05 , 185.64) \\ 
  20 & 300.72 & (291.72 , 309.99) & 304.58 & (297.40 , 311.94) & 255.25 & (246.66 , 264.14) & 239.05 & (232.32 , 245.97) \\ 
  30 & 339.98 & (332.69 , 347.42) & 372.71 & (365.91 , 379.64) & 280.49 & (272.82 , 288.39) & 284.33 & (277.68 , 291.14) \\ 
  40 & 371.65 & (364.63 , 378.81) & 404.92 & (398.26 , 411.70) & 289.70 & (282.62 , 296.95) & 291.85 & (285.77 , 298.05) \\ 
  50 & 406.28 & (398.43 , 414.29) & 439.92 & (432.44 , 447.53) & 299.20 & (291.71 , 306.90) & 299.56 & (293.11 , 306.16) \\ 
  60 & 444.14 & (434.12 , 454.39) & 477.94 & (468.53 , 487.55) & 309.02 & (300.05 , 318.26) & 307.48 & (299.70 , 315.47) \\ 
  70 & 485.52 & (472.13 , 499.30) & 519.25 & (506.89 , 531.92) & 319.17 & (307.94 , 330.80) & 315.61 & (305.83 , 325.71) \\ 
  80 & 530.76 & (512.94 , 549.21) & 564.13 & (547.94 , 580.80) & 329.64 & (315.62 , 344.28) & 323.96 & (311.74 , 336.66) \\ 
  90 & 505.34 & (477.57 , 534.72) & 533.79 & (506.27 , 562.81) & 320.11 & (299.52 , 342.12) & 312.65 & (293.61 , 332.92) \\ 
  100 & 419.04 & (362.32 , 484.63) & 439.90 & (380.85 , 508.10) & 292.28 & (246.86 , 346.06) & 283.71 & (239.92 , 335.49) \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[]
\centering
\caption{Predicted values with 95\% CI for female patients with acuity level 1,4 and 5 (symptom type=4)}
\label{my-label}
\begin{tabular}{|l|ll|ll|ll|ll|}
\hline
\multicolumn{1}{|c|}{\textbf{Day}}  & \multicolumn{4}{c|}{\textbf{Weekday}}                                                               & \multicolumn{4}{c|}{\textbf{Weekend}}                                                               \\ \hline
\multicolumn{1}{|c|}{\textbf{Time}} & \multicolumn{2}{c|}{\textbf{{[}12AM-11:59AM{]}}} & \multicolumn{2}{c|}{\textbf{{[}12PM-11:59PM{]}}} & \multicolumn{2}{c|}{\textbf{{[}12AM-11:59AM{]}}} & \multicolumn{2}{c|}{\textbf{{[}12PM-11:59PM{]}}} \\ \hline
Age                                 & $\widehat{los}$               & 95\% CI              & $\widehat{los}$               & 95\% CI              & $\widehat{los}$               & 95\% CI              & $\widehat{los}$               & 95\% CI              \\ \hline  
  10 & 160.89 & (142.67 , 181.44) & 138.25 & (125.28 , 152.55) & 156.28 & (138.48 , 176.37) & 124.17 & (112.33 , 137.25) \\ 
  20 & 166.57 & (160.00 , 173.40) & 168.71 & (163.10 , 174.51) & 161.80 & (154.68 , 169.24) & 151.53 & (145.69 , 157.60) \\ 
  30 & 175.80 & (170.49 , 181.27) & 192.73 & (187.33 , 198.27) & 165.98 & (159.80 , 172.41) & 168.25 & (162.50 , 174.21) \\ 
  40 & 189.15 & (184.19 , 194.25) & 206.09 & (201.01 , 211.29) & 168.73 & (162.98 , 174.69) & 169.98 & (164.64 , 175.50) \\ 
  50 & 203.52 & (197.60 , 209.63) & 220.38 & (214.25 , 226.68) & 171.53 & (165.17 , 178.12) & 171.73 & (165.78 , 177.90) \\ 
  60 & 218.99 & (210.83 , 227.46) & 235.65 & (227.22 , 244.40) & 174.37 & (166.55 , 182.55) & 173.50 & (166.12 , 181.20) \\ 
  70 & 235.62 & (224.34 , 247.47) & 251.99 & (240.40 , 264.14) & 177.25 & (167.45 , 187.64) & 175.28 & (166.03 , 185.05) \\ 
  80 & 253.52 & (238.42 , 269.59) & 269.46 & (254.06 , 285.80) & 180.19 & (168.09 , 193.17) & 177.09 & (165.70 , 189.25) \\ 
  90 & 211.03 & (196.20 , 226.98) & 222.91 & (207.96 , 238.94) & 152.98 & (141.12 , 165.84) & 149.42 & (138.34 , 161.38) \\ 
 100 & 135.89 & (116.10 , 159.05) & 142.65 & (122.16 , 166.58) & 108.47 & (91.24 , 128.96)  & 105.29 & (88.76 , 124.89) \\ 
   \hline
\end{tabular}
\end{table}

Through Table 1 and Table 2 above, we could conclude that, female patients with acuity level 2 or 3 will have longer predicted length of stay in all age groups compared with other acuity levels regardless of arrival time. And if female patients arrive on weekday they will have longer predicted length of stay regardless other predictors. The relationship between predicted length of stay and arrival time varies with age. And the 95\% confidence intervals are narrow enough so that the variance of predicted values are plausible.

```{r message=F,warning=F, echo=F,fig.height=3.5}
# predicted values
#op <- par(cex=.5)
#plot(data$age,data$los,cex=0.3,col=rgb(0.5,0.5,0.5,0.5),pch=16,xlab="age",ylab="los",main="predicted values of the poisson regression model")
#lines(ksmooth(sort(data$age),pred[order(data$age)],kernel="normal",bandwidth = 10),col=2,lwd=2)
#lines(ksmooth(sort(data$age),output$uppery[order(data$age)],kernel="normal",bandwidth = 10),col=4,lwd=2)
#points(sort(data$age),pred[order(data$age)],col=3,type="l",lwd=2)
#par(op)

 #fig.predict<- ggplot(output,aes(age,los))+ theme_bw()+geom_jitter(aes(y=los),size=0.2,alpha=0.5,colour="grey")
  
#geom_jitter(aes(y=los),size=0.1,alpha=0.5,colour="grey")+
#fig.ci<-ggplot(output,aes(x=age))+  theme_bw()+ geom_line(aes(y=uppery),colour="orange",size=0.4,alpha=0.7)+ geom_line(aes(y=lowery),,colour="orange",size=0.4,alpha=0.7)+geom_line(aes(y=yhat),size=0.4,alpha=1,colour="sky blue")+scale_color_discrete(labels=c("a","b"))+theme(legend.position="bottom")+ylab("los")+ggtitle("Fig 7: predicted values with 95% confidence interval")
#fig.ci
```

In conclusion, when controlled for other predictors, the predicted length of stay gets longer with increase of age when people are younger than 85 years old, while it will get shorter with increase of age when people are older than 85. Patients in same condition will have longer predicted length of stay if they arrive on weekday, with acuity level 2 or 3, or have high level symptom type (type 6, 5, 4). The relationship between age and predicted length of stay is also modified by acuity level, gender, arrival day and time.

\newpage

# Supplement

## I. Symptom Types

Symptom types category is shown in Table 3 below.

\begin{table}[]
\centering
\caption{Symptom Types}
\begin{tabular}{|lll|lll|lll|}
\hline
key        & meanlos     & type & key        & meanlos     & type & key           & meanlos     & type \\ \hline
aaa        & 58          & 1    & suture     & 69 & 1    & cough         & 378 & 4    \\ \hline
abd        & 83 & 1    & sz         & 70          & 1    & depress       & 354 & 4    \\ \hline
abscess    & 13          & 1    & test       & 22          & 1    & flank         & 393 & 4    \\ \hline
all over   & 28          & 1    & thigh      & 38          & 1    & flu           & 370 & 4    \\ \hline
aneurysm   & 68          & 1    & to or      & 49          & 1    & groin         & 342 & 4    \\ \hline
asthma     & 14        & 1    & toe        & 32          & 1    & gu            & 399    & 4    \\ \hline
bleeding   & 5           & 1    & tooth      & 42          & 1    & ha            & 301 & 4    \\ \hline
bp         & 67 & 1    & unknown    & 35          & 1    & head          & 300 & 4    \\ \hline
c-sec comp & 11          & 1    & a\&d       & 117      & 2    & headache      & 385 & 4    \\ \hline
check      & 12          & 1    & bness      & 134       & 2    & hematuria     & 369 & 4    \\ \hline
clot       & 60          & 1    & cancer     & 118 & 2    & hip           & 347 & 4    \\ \hline
confusion  & 75          & 1    & cardiac    & 144      & 2    & jaundice      & 303         & 4    \\ \hline
cramp      & 10          & 1    & cath       & 115         & 2    & palp          & 348 & 4    \\ \hline
dehydr     & 40          & 1    & congestion & 189 & 2    & phylaxis      & 304   & 4    \\ \hline
deliver    & 21          & 1    & cp         & 135         & 2    & preg          & 351 & 4    \\ \hline
detox      & 86          & 1    & cva        & 169         & 2    & problem       & 326  & 4    \\ \hline
disorie    & 43          & 1    & dynia      & 159 & 2    & rib           & 315  & 4    \\ \hline
etoh       & 38          & 1    & ear        & 133 & 2    & scan          & 313         & 4    \\ \hline
eval       & 100 & 1    & hand       & 193 & 2    & skin          & 330  & 4    \\ \hline
faint      & 45          & 1    & lab        & 178 & 2    & abnormal      & 467 & 5    \\ \hline
finger     & 87          & 1    & op comp    & 105 & 2    & ad            & 435  & 5    \\ \hline
flash      & 47          & 1    & psych      & 127         & 2    & complication  & 430 & 5    \\ \hline
follow     & 24          & 1    & refill     & 108  & 2    & diz           & 411 & 5    \\ \hline
foot       & 73          & 1    & s/p        & 185         & 2    & general       & 488 & 5    \\ \hline
gen ill    & 58          & 1    & sah        & 116         & 2    & hemoptysis    & 497 & 5    \\ \hline
heart      & 78          & 1    & snake      & 178 & 2    & hypoglycemia  & 490  & 5    \\ \hline
infect     & 5           & 1    & wrist      & 174  & 2    & kidney        & 417 & 5    \\ \hline
liver      & 12          & 1    & anxi       & 278 & 3    & muscle        & 437 & 5    \\ \hline
med        & 16          & 1    & arm        & 279 & 3    & pe            & 479 & 5    \\ \hline
migra      & 9           & 1    & back       & 295 & 3    & seizure       & 408 & 5    \\ \hline
mouth      & 81          & 1    & dtoh       & 207         & 3    & si            & 465 & 5    \\ \hline
nausea     & 40     & 1    & dysuria    & 290 & 3    & site          & 458 & 5    \\ \hline
needs      & 73          & 1    & elbow      & 229 & 3    & stool         & 466  & 5    \\ \hline
nose       & 76          & 1    & epistaxis  & 289 & 3    & tachycardia   & 492 & 5    \\ \hline
passed out & 9           & 1    & eye        & 226 & 3    & vertigo       & 465 & 5    \\ \hline
picu       & 56          & 1    & feet       & 201         & 3    & wheezing      & 491 & 5    \\ \hline
poison     & 60      & 1    & insom      & 279      & 3    & ams           & 591         & 6    \\ \hline
pressure   & 75          & 1    & knee       & 234 & 3    & breath        & 553 & 6    \\ \hline
rabies     & 70          & 1    & leg        & 297 & 3    & chestpain     & 527 & 6    \\ \hline
rectal     & 14          & 1    & lle        & 245 & 3    & chill         & 557         & 6    \\ \hline
removal    & 33          & 1    & n/v        & 247         & 3    & clearance     & 566         & 6    \\ \hline
respira    & 78 & 1    & neck       & 263 & 3    & constipation  & 510 & 6    \\ \hline
results    & 6           & 1    & obs        & 210         & 3    & diarr         & 509 & 6    \\ \hline
rx         & 42          & 1    & pain       & 283 & 3    & dyspnea       & 524 & 6    \\ \hline
sa         & 87          & 1    & rash       & 200  & 3    & extremity     & 501 & 6    \\ \hline
scc        & 44          & 1    & shoulder   & 229 & 3    & fever         & 646 & 6    \\ \hline
shiver     & 81          & 1    & sore       & 205   & 3    & gi            & 505 & 6    \\ \hline
shot       & 22          & 1    & throat     & 273   & 3    & hyperglycemia & 510 & 6    \\ \hline
sinu       & 35        & 1    & transfer   & 233         & 3    & ingestion     & 541       & 6    \\ \hline
sob        & 90 & 1    & urticaria  & 229 & 3    & lung          & 600      & 6    \\ \hline
std        & 24          & 1    & voice      & 239 & 3    & melena        & 613 & 6    \\ \hline
stitch     & 76          & 1    &            &             &      & mi            & 530  & 6    \\ \hline
stoma      & 42 & 1    &            &             &      & null          & 1133        & 6    \\ \hline
sugar      & 76          & 1    &            &             &      & stridor       & 531 & 6    \\ \hline
surg       & 68          & 1    &            &             &      & thrive        & 1022        & 6    \\ \hline
           &             &      &            &             &      & weak          & 578 & 6    \\ \hline
           &             &      &            &             &      & weight        & 584 & 6    \\ \hline
\end{tabular}
\end{table}


## II. Assumptions Checking

###1. Independence assumption
 
```{r results='hide',message=F,warning=F, echo=F, fig.height=4.3,fig.width=4.8,dev="png",dpi=300}
pred<-predict.glm(fit,type="response")
resi<-data$los-pred
presi<-(data$los-pred)/sqrt(pred+pred^2*fit$theta)
#plot(pred,presi)
#abline(h=0,col=2,lwd=2)
prepresi<-rep(NA,length(presi)+1)
prepresi[2:length(prepresi)]<-presi
a<-prepresi[2:length(presi)]
b<-presi[2:length(presi)]
plot(a,b,main="Fig S1: correlation betwen two patients",xlab="los for patient (i-1)",ylab="los for patient i",col=rgb(0.2,0.2,0.3,0.5),cex=0.5,pch=16)
lines(ksmooth(a,b,kernel="normal",bandwidth=5),col="red",lwd=2)
```


### 2. Average Standardized residual


```{r results='hide',message=F,warning=F, echo=F, fig.height=4.5,fig.width=5,dev="png",dpi=300}
# average resis = 0
data$pred<-pred
data$presi<-presi
binsize<-80
meanpres<-rep(NA,binsize)
meanpred<-rep(NA,binsize)
for (i in 1:binsize){
  if(i!=binsize){
    subdata<-data[data$pred>=quantile(pred,probs=(i-1)/binsize) &
                    data$pred<quantile(pred,probs=i/binsize),]
  }
  else{
    subdata<-data[data$pred>=quantile(pred,probs=(i-1)/binsize) &
                    data$pred<=quantile(pred,probs=i/binsize),]
  }
  meanpres[i]<-mean(subdata$pres)
  meanpred[i]<-mean(subdata$pred)
}
plot(pred,presi,col=rgb(0.2,0.2,0.3,0.5),cex=0.5,pch=16,xlab="predicted values",ylab="Standardized residuals",main="Fig S2")
points(meanpred,meanpres,col=rgb(0.8,0.2,0.4,0.9),cex=0.8,pch=16)
lines(ksmooth(meanpred,meanpres,kernel="normal",bandwidth = 10),col="red",lwd=2)
lines(ksmooth(pred,presi,kernel="normal",bandwidth = 10),lty=3)
op <- par(cex=0.7)
legend("topleft",legend=c("average standardized residuals","smooth curve for average residuals","standardized residuals","smooth curve for all residual"),
       col=c(rgb(0.8,0.2,0.4,0.9),"red",rgb(0.2,0.2,0.3,0.5),"black"),pch=c(16,NA,16,NA),lty=c(NA,1,NA,2),lwd=c(NA,2,NA,1))
par(op)
```


### 3. Highly influency points


```{r warning=F, echo=F}
# highly influetial points
dfit<-dffits(fit)
#plot(pred,dfit,ylab="dffits",xlab="predicted values",col=rgb(0.2,0.2,0.3,0.6),cex=0.7,pch=16,main="Fig S3: Dffits plot")
#abline(h=0.17,col=2)
#abline(h=-0.17,col=2)
```

```{r warning=F, echo=F, fig.height=4.3,fig.width=4.8,dev="png",dpi=300}
dropdata<-data[abs(dfit)<=0.17,]
#nrow(data)-nrow(dropdata) #59
fitdrop<-glm.nb(los~age*(acu23+prenoon)+age_sp1*(acu23+weekday+prenoon)+age_sp2*(acu23+gender+weekday)+acu23*weekday+ weekday*prenoon + factor(type), data=dropdata)
preddrop<-predict.glm(fitdrop,type="response")
```

```{r echo=F,fig.height=4.3,fig.width=4.8,dev="png",dpi=300}
plot(data$age,pred,cex=0,main="Fig S4: smooth curve of predicated values",
     ylab="predicted daily number of deaths",xlab="day")
lines(ksmooth(data$age,pred,kernel="normal",bandwidth = 10),col="black",lwd=2)
lines(ksmooth(dropdata$age,preddrop,kernel="normal",bandwidth = 10),col="red",lwd=2)
legend("topleft",legend=c("before dropping outliers","after dropping outliers"),
       col=c("black","red"),lwd=2)
```


### 4. Cross-validation

```{r echo=F,fig.height=3.6 ,dev="png",dpi=300}
splitdata<-split(data,sample(rep(1:11,nrow(data)/11)))
cv.resi<-NA
cv.stresi<-NA
pred.cv<-NA
for (i in 1:11){
  test<-splitdata[[i]]
  train<-data[-c(as.numeric(row.names(test))),]
  fitnew1<-glm.nb(los~age*(acu23+prenoon)+age_sp1*(acu23+weekday+prenoon)+age_sp2*(acu23+gender+weekday)+acu23*weekday+ weekday*prenoon + factor(type), data=train)
  predrate1<-predict.glm(fitnew1,newdata=test,type="response")
  predrate.cv<-predrate1
  obserate.cv<-test$los
  pred.cv<-c(pred.cv,predrate.cv)
  cv.resi<-c(cv.resi,obserate.cv-predrate.cv)
  cv.stresi<-c(cv.stresi,(obserate.cv-predrate.cv)/sqrt(predrate.cv+predrate.cv^2*fitnew1$theta))
}
cv.resi<-cv.resi[complete.cases(cv.resi)]
par(mfrow=c(1,2))
plot(pred,presi,col=rgb(0.2,0.2,0.3,0.6),cex=0.7,pch=16,main="fitted model without cross-validation",xlab="predicted values",ylab="standardized residul")  #original
plot(pred.cv,cv.stresi,col=rgb(0.2,0.2,0.3,0.6),cex=0.7,pch=16,main="11-fold cross-validation",xlab="predicted values",ylab="standardized residul") #cv
```
